---
title: "Getting started with disprr"
author: "Michal Pierzgalski"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with disprr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 96
)
```

## What `disprr` does

`disprr` is a toolkit for studying how **proportional representation (PR)
electoral systems** translate votes into seats. It has three layers:

1. **Apportionment** -- given a vector of votes, allocate a fixed number of
   seats to parties using one of eleven divisor methods or the Hamilton-Hare
   largest-remainder method.
2. **Simulation** -- generate many synthetic elections from a probability
   distribution and run them through an apportionment method.
3. **Disproportionality analysis** -- summarise how far the seat distribution
   departs from the vote distribution, both per party and with aggregate
   indexes, and plot the result against district magnitude.

This guide walks through all three layers.

> **A note on the numbers in this guide.** Every vote count shown below is
> *illustrative and synthetic* -- it is invented for teaching purposes and
> does not describe any real election. Likewise, the simulation sections
> draw pseudo-random data; the figures they produce are generated when this
> vignette is built, not copied from any data source.

```{r library}
library(disprr)
```

## 1. Apportioning seats

The two entry points are `divisorMethods()` and `LR_Hamilton()`. Both take a
vector of party labels, a vector of votes, and a seat total, and both return
a data frame with one row per party.

```{r apportion-basic}
parties <- c("Liberals", "Centre", "Greens", "Nationals", "Others")
votes   <- c(100000, 80000, 30000, 20000, 10000)

divisorMethods(parties = parties, votes = votes, seats = 10, method = "dh")
```

The output columns are:

- `Party`, `Votes`, `VoteShare` -- the inputs and each party's share of all
  votes cast.
- `Seats`, `SeatShare` -- the allocation and each party's share of seats.

### Choosing a method

The `method` argument selects the divisor sequence. The same votes apportioned
by Sainte-Laguë give a more proportional result than D'Hondt:

```{r apportion-sl}
divisorMethods(parties = parties, votes = votes, seats = 10, method = "sl")
```

The full set of divisor methods is:

| Code          | Method                  | Divisor sequence            |
|:--------------|:------------------------|:----------------------------|
| `"dh"`        | D'Hondt                 | 1, 2, 3, 4, ...             |
| `"sl"`        | Sainte-Laguë | 1, 3, 5, 7, ...             |
| `"msl"`       | Modified Sainte-Laguë | 1.4, 3, 5, 7, ...           |
| `"danish"`    | Danish                  | 1, 4, 7, 10, ...            |
| `"hsl"`       | Hungarian Sainte-Laguë | 1.5, 3, 5, 7, ...           |
| `"imperiali"` | Imperiali (divisor)     | 1, 1.5, 2, 2.5, ...         |
| `"hh"`        | Huntington-Hill         | 0, $\sqrt{2}$, $\sqrt{6}$, ... |
| `"wb"`        | Webster                 | 0.5, 1.5, 2.5, ...          |
| `"jef"`       | Jefferson               | 1, 2, 3, 4, ...             |
| `"ad"`        | Adams                   | 0, 1, 2, 3, ...             |
| `"hb"`        | Hagenbach-Bischoff      | 1, 2, 3, 4, ...             |

Several of these are mathematically identical -- they assign the same seats
for any input -- because their divisor sequences differ only by a constant
factor, which does not change the ranking of quotients:

```{r equivalences}
seats_dh  <- divisorMethods(parties, votes, seats = 10, method = "dh")$Seats
seats_jef <- divisorMethods(parties, votes, seats = 10, method = "jef")$Seats
seats_hb  <- divisorMethods(parties, votes, seats = 10, method = "hb")$Seats
identical(seats_dh, seats_jef) && identical(seats_dh, seats_hb)

seats_sl <- divisorMethods(parties, votes, seats = 10, method = "sl")$Seats
seats_wb <- divisorMethods(parties, votes, seats = 10, method = "wb")$Seats
identical(seats_sl, seats_wb)
```

So D'Hondt = Jefferson = Hagenbach-Bischoff, and Sainte-Laguë = Webster.

Two methods -- Huntington-Hill and Adams -- have a first divisor of `0`. Every
party with positive votes therefore has an infinite first quotient, which
guarantees it at least one seat whenever there are at least as many seats as
parties. Huntington-Hill (the "method of equal proportions") is the method
used to apportion seats in the U.S. House of Representatives, adopted in the
1941 Apportionment Act.

### The Hamilton-Hare largest-remainder method

`LR_Hamilton()` is the quota-based alternative. It gives each party the whole
number of quotas it earns, then hands the leftover seats to the parties with
the largest fractional remainders:

```{r hamilton}
LR_Hamilton(parties = parties, votes = votes, seats = 10)
```

### Electoral thresholds

Both functions accept a `threshold`: a vote-share cut-off in `[0, 1]`. Parties
*strictly below* the threshold are excluded before seats are allocated; a
party whose share exactly equals the threshold is kept.

With these illustrative votes, "Others" holds `r round(10000 / 240000 * 100, 1)`%
of the vote, so a 5% threshold removes it:

```{r threshold}
divisorMethods(parties, votes, seats = 10, method = "dh", threshold = 0.05)
```

## 2. Simulating elections

Real elections give one observation at a time. To study the *behaviour* of an
apportionment rule we simulate many elections and look at the distribution of
outcomes.

`sampleElectionData()` draws synthetic votes; `simulate_E()` wraps it together
with apportionment and disproportionality scoring. A single call to
`simulate_E()` runs `ne` independent elections with the same settings:

```{r simulate-e}
sim <- simulate_E(
  seed = 1, dist = "lnorm",
  np = 5, nd = 1, ne = 50,
  mean = 10, sd = 1.2, rate = 1 / 25000, max = 100000,
  TS = 11, formula = "dh", formula_dist = "hh",
  threshold = 0, threshold_country = 0
)
names(sim)
```

The `Disproportionality_per_elec` element holds one row of aggregate indexes
per simulated election:

```{r simulate-e-disp}
head(sim$Disproportionality_per_elec)
```

and `Summary` gives descriptive statistics of those indexes across all `ne`
elections:

```{r simulate-e-summary}
sim$Summary
```

Key arguments:

- `dist` -- the vote-generating distribution. Five options are available, in
  two families. **Independent count distributions** (`"uniform"`,
  `"lnorm"` (log-normal), `"exp"` (exponential)) draw each party's vote
  count independently. **Compositional distributions** draw a vector of
  vote shares on the simplex and scale it by `votes_per_district`:
  `"dirichlet"` is an asymmetric Dirichlet calibrated via the
  Taagepera-Allik rule -- a realistic party-system prior, following
  Cohen & Hanretty (2024); `"uniform_simplex"` is the symmetric
  Dirichlet(1, ..., 1), the Pukelsheim benchmark in which every vector
  of vote shares is equally likely.
- `np`, `nd`, `ne` -- number of parties, districts, and elections.
- `TS` -- total seats; `formula` -- the apportionment method;
  `formula_dist` -- the method used to share `TS` seats among districts.
- `phi`, `votes_per_district` -- precision and total electorate for the
  Dirichlet distributions.
- `threshold`, `threshold_country` -- district- and country-level
  vote-share thresholds.

A quick look at the Taagepera-Allik mean vector for a 5-party system:

```{r taagepera-example}
taagepera_allik(5)
```

A Dirichlet draw with that mean and precision `phi = 30` is then a noisy
realisation of the same structure:

```{r dirichlet-example}
disp_dir <- simulate_E(
  seed = 1, dist = "dirichlet",
  np = 5, nd = 1, ne = 50,
  phi = 30, votes_per_district = 100000,
  TS = 11, formula = "dh"
)
head(disp_dir$Disproportionality_per_elec)
```

Note that the log-normal parameters (`mean`, `sd`, `max`) are not needed
here -- the Dirichlet models only require `phi` and `votes_per_district`.

## 3. Disproportionality across district magnitude

District magnitude (the number of seats per district) is the single strongest
driver of disproportionality. `simulate_Disp()` sweeps a range of seat totals
and records per-party seat excess at each one.

```{r simulate-disp}
bias <- simulate_Disp(
  seed = 1, dist = "lnorm",
  np = 5, nd = 1, ne = 30,
  formula = "dh",
  minTS = 3, maxTS = 15, jump = 2
)
```

`plot_Disp()` turns the result into a list of diagnostic `ggplot` objects.
The second plot shows each party's seat-share-minus-vote-share (`SE2_i`),
faceted by party, with one box per district magnitude:

```{r plot-disp, fig.height = 5}
disp_plots <- plot_Disp(bias)
disp_plots[[2]]
```

As district magnitude rises (lighter boxes), the boxes move toward zero: more
seats per district means a fairer translation of votes into seats.

## 4. Comparing apportionment methods

`Disp2()` runs the same sweep for several methods at once and collects the
aggregate indexes. It optionally fits an exponential decay model
$\mathrm{GHI} = C \cdot e^{\alpha \cdot DM}$ to the Gallagher index.

```{r disp2}
comparison <- Disp2(
  seed = 1, np = 5, ne = 40,
  minTS = 3, maxTS = 15, jump = 2,
  methods = c("dh", "sl", "hamilton", "hh")
)
```

`plot_Disp2()` plots the indexes against district magnitude, one facet per
method. Its `methods` argument expects the short *labels* (`"DH"`, `"SL"`,
`"H"`, `"HH"`, ...), not the lower-case codes:

```{r plot-disp2, fig.height = 5}
gg <- plot_Disp2(comparison, methods = c("DH", "SL", "H", "HH"))
gg$plot_GHI
```

The Gallagher index falls steeply as district magnitude grows, and the four
methods converge: at large magnitudes the choice of formula barely matters.

## 5. The disproportionality indexes

For party $i$ with vote share $v_i$ and seat share $s_i$, `disprr` reports
four aggregate indexes per election.

The **Gallagher index** (least-squares index) takes the root of the mean
squared deviation, so a few large deviations weigh more heavily than many
small ones:

$$
\mathrm{GHI} = \sqrt{\tfrac{1}{2} \sum_i (s_i - v_i)^2}
$$

The **Loosemore-Hanby index** is the total over-representation (equivalently,
total under-representation):

$$
\mathrm{LHI} = \tfrac{1}{2} \sum_i |s_i - v_i|
$$

The **Sainte-Laguë index** weights each squared deviation by the party's
vote share, making it sensitive to small parties:

$$
\mathrm{SLI} = \sum_i \frac{(s_i - v_i)^2}{v_i}
$$

The **effective number of parliamentary parties** summarises seat
concentration:

$$
\mathrm{ENPP} = \frac{1}{\sum_i s_i^2}
$$

All four are computed from full-precision vote and seat shares. They are on a
0--1 (proportion) scale, not a 0--100 (percentage) scale: a Gallagher index of
`0.05` corresponds to the value often reported as "5".

## Where to go next

- `?divisorMethods`, `?LR_Hamilton` -- apportionment, with worked examples.
- `?simulate_E`, `?simulate_Disp`, `?Disp2` -- the simulation engine.
- `?plot_Disp`, `?plot_Disp2` -- the diagnostic plots.

## References

Balinski, M. L., & Young, H. P. (2001). *Fair Representation: Meeting the
Ideal of One Man, One Vote* (2nd ed.). Brookings Institution Press.

Cohen, D., & Hanretty, C. (2024). Simulating Party Shares. *Political
Analysis*, 32(1), 140--147.

Gallagher, M. (1991). Proportionality, disproportionality and electoral
systems. *Electoral Studies*, 10(1), 33--51.
<https://doi.org/10.1016/0261-3794(91)90004-C>

Schuster, K., Pukelsheim, F., Drton, M., & Draper, N. R. (2003). Seat biases
of apportionment methods for proportional representation. *Electoral
Studies*, 22(4), 651--676.

Taagepera, R., & Allik, M. (2006). Seat Share Distribution of Parties:
Models and Empirical Patterns. *Electoral Studies*, 25(4), 696--713.
